---
title: "Getting started with CanPAS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CanPAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4, warning = FALSE, message = FALSE)
set.seed(20260930)
```

```{r}
library(CanPAS)
```

## What the package is for

CanPAS runs one survival analysis over many public cancer cohorts and one
uniform data schema. A cohort is identified by its accession (`GSE31210`,
`TCGA-LUAD`, `CGGA_693`, ...); an endpoint is requested by **family** (`OS`,
`DSS`, `DFS`, `PFS`, `MFS`) and the concrete endpoint token a cohort actually
reports (`RFS`, `PFI`, `DFI`, ...) is resolved per cohort and returned with every
result. This vignette is deliberately **offline**: it uses the shipped catalog
and small synthetic cohorts so that every number below is reproducible without a
mirror, a network connection or a download.

## The catalog

`dataset_info` is one row per curated cohort, with the sample-size columns and
the endpoint-family mapping.

```{r}
di <- dataset_info
dim(di)
head(di[, c("Accession", "Type", "N", "EndpointFamilies")], 4)
```

Which families a cohort can be analysed on, and with which token:

```{r}
endpoint_options("TCGA-LGG")
endpoint_resolve("GSE31210", "DFS")   # that cohort reports RFS, not DFS
endpoint_resolve("TCGA-LUAD", "PFS")  # a TCGA-derived endpoint
```

## Getting real data

On a machine that can reach the CanPAS mirror (or with the local artefact tree
present, via `CPAS_DATA_ROOT`), expression and survival are fetched and merged in
one call. These chunks are not evaluated here because they need the mirror.

```{r eval = FALSE}
d <- cohort_merged("GSE31210", c("TP53", "GAPDH"), type = "DFS", clin = TRUE)
colnames(d)
```

```{r eval = FALSE}
# a signature instead of one gene: any R expression on the gene columns
d <- cohort_merged("GSE31210", NULL, type = "RFS", signature = "0.5*TP53+0.5*GAPDH")
```

## An end-to-end analysis, offline

We build a small cohort that has exactly the merged schema (`ID`, `<token>_time`,
`<token>_status`, one column per marker) and run the same functions the package
runs on real data.

```{r}
mk_cohort <- function(n, prefix, seed) {
  set.seed(seed)
  x <- rnorm(n, 10, 1)
  hr <- exp(0.35 * (x - 10))
  data.frame(ID = sprintf("%s%03d", prefix, seq_len(n)),
             OS_time = round(rexp(n, hr / 8) + 0.05, 3),
             OS_status = rbinom(n, 1, 0.65),
             TP53 = x,
             GAPDH = rnorm(n, 12, 1),
             age = round(runif(n, 35, 82)),
             stage = factor(sample(c("I", "II", "III"), n, TRUE,
                                   prob = c(0.4, 0.35, 0.25))))
}
d1 <- mk_cohort(220, "P", 1)
d2 <- mk_cohort(180, "Q", 2)
```

Univariable Cox, one model per covariate:

```{r}
u <- COX_analysis(d1, type = "OS", cont_Variates = c("TP53", "age"),
                  cate_Variates = "stage", method = "uni")
u$results_table[, c("Variates", "Level", "N", "HR", "HR95L", "HR95H", "Pvalue")]
```

A multivariable model. Since this version the default is the **fail-safe**
(`auto_repair = FALSE`): if a covariate cannot be co-estimated, the model is
*refused* with its reasons instead of being quietly reduced.

```{r}
m <- COX_analysis(d1, type = "OS", cont_Variates = c("TP53", "age"),
                  cate_Variates = "stage", method = "multi")
m$estimable
m$results_table[, c("Variates", "Level", "HR", "Pvalue")]
```

```{r}
d1$flat <- 1                       # a covariate with no variation at all
bad <- COX_analysis(d1, type = "OS", cont_Variates = c("TP53", "flat"),
                    method = "multi")
bad$estimable
bad$offending_terms
bad$reasons
```

The refusal is a value, not an exception, so a script does not have to wrap the
call in `tryCatch()`. Ask for the old behaviour explicitly and every modification
is recorded:

```{r}
fixed <- COX_analysis(d1, type = "OS", cont_Variates = c("TP53", "flat"),
                      method = "multi", auto_repair = TRUE)
fixed$metadata$dropped_covariates[, c("variable", "reason")]
```

Kaplan-Meier with a cut-point rule:

```{r, fig.height = 4.5}
km <- plot_km(d1, type = "OS", marker = "TP53", cutpoint = "median")
km$logrank_p
```

Time-dependent ROC at one, three and five years (`plot_roc()` returns one
ggplot per time point, with the estimated AUC attached as the `"AUC"`
attribute):

```{r, fig.height = 4.5}
roc <- lapply(c(1, 3, 5), function(t)
  plot_roc(d1, type = "OS", marker = "TP53", predict.time = t))
round(vapply(roc, function(p) attr(p, "AUC"), numeric(1)), 3)
roc[[1]]
```

## Two-stage meta-analysis across cohorts

The same marker in several cohorts is pooled in one call. This version defaults
to `method = "REML"`; `"DL"` reproduces the previous estimator and `"HK"` gives
the Hartung-Knapp-Sidik-Jonkman interval. One call always reports the token each
cohort contributed, the pooling class of that token and the manifest of the run.

```{r}
mt <- cpas_meta(c("P", "Q"), marker = "TP53", type = "OS",
                merged = list(P = d1, Q = d2))
mt$per_dataset[, c("dataset", "endpoint", "pooling_class", "n", "events", "HR")]
mt$pooled[, c("method", "k", "HR", "lower", "upper", "p", "tau2", "I2",
              "pi_lower", "pi_upper", "pi_alt_lower", "pi_alt_upper")]
```

## The manifest

Every analysis result carries a manifest: which cohorts, which token, which
pooling classes, what was dropped and why, the meta-analysis method, tau^2, I^2,
both prediction intervals, the versions and the timestamp. It prints and it
converts to a table.

```{r}
cpas_manifest(mt)
```

```{r}
head(as.data.frame(cpas_manifest(mt)), 8)
```

## The Shiny application

The bundled application drives the same exported functions.

```{r eval = FALSE}
run_cpas_app()
```

## Where to go next

* `vignette("endpoints-and-pooling", package = "CanPAS")` covers the endpoint
  semantics layer, the two pooling modes, shared-patient overlap and the
  manifest in detail.
* `?cpas_meta`, `?COX_analysis`, `?cpas_km_pooled`, `?plot_meta_forest` are the
  reference pages for the analysis functions.
