---
title: "Endpoints, pooling classes and shared-patient overlap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Endpoints, pooling classes and shared-patient overlap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4, warning = FALSE, message = FALSE)
set.seed(20260930)
```

```{r}
library(CanPAS)
```

## Why endpoints need a semantics layer

Different cohorts call the same clinical question by different names. "How long
until the disease comes back?" may be deposited as `DFS`, `RFS`, `EFS` or `DFI`;
"how long until progression?" as `PFS` or the TCGA-derived `PFI`. CanPAS groups
those tokens into five **pooling families** (`OS`, `DSS`, `DFS`, `PFS`, `MFS`),
resolves the token each cohort actually provides, and records what that token
means. Nothing is pooled across families.

```{r}
endpoint_family("RFS")     # a token -> its family
endpoint_family("DRFS")
endpoint_options("GSE31210")
```

The registry behind that is the **endpoint semantics table**: one row per cohort
and family, with the token, the role of the token, what the deposit states about
the event definition / time origin / censoring / competing events, the artefact
the entry was read from, and - the column the analysis uses - the **pooling
class**.

```{r}
es <- endpoint_semantics(family = "DFS")
es[es$Accession %in% c("GSE31210", "GSE13507", "TCGA-LUAD"),
   c("Accession", "Family", "Token", "TokenRole", "PoolingClass", "Evidence")]
```

```{r}
table(endpoint_semantics()$PoolingClass)
```

The vocabulary is frozen and has five values:

| class | meaning |
|---|---|
| `Exact-equivalent` | the token **is** the family's canonical definition (OS in OS, DFS in DFS, ...) |
| `Clinically-related` | a different token pooled into the family by the documented rule (RFS/EFS/DFI in DFS, CSS/BCSS in DSS, PFI in PFS, DRFS in MFS) |
| `Not-poolable` | the cohort reports an endpoint in this family that cannot be pooled |
| `Unknown` | the deposit does not state what the endpoint means |
| `Absent` | the cohort has no endpoint of this family at all - the absence of an endpoint, **not** a pooling verdict |

```{r}
endpoint_pooling_class("GSE31210", "DFS")        # RFS: related, not identical
endpoint_pooling_class("GSE4922_GPL96", "DFS")   # DFS itself: exact
endpoint_pooling_class("GSE13507", "DFS")        # no DFS endpoint at all -> Absent
```

`endpoint_pooling_class()` is one cell of `endpoint_semantics()`. If the
companion table is not shipped in a build, both fall back to the package's
documented token -> family map, which is the same rule the table records for the
exact/related split; a family the catalog does not annotate still comes back as
`Absent`.

## Two pooling modes

`cpas_meta()` has two modes and `"family"` is the documented default:

* `pooling = "family"` (default) pools `Exact-equivalent` **and**
  `Clinically-related` rows of one family. This is what makes a cross-token claim
  possible at all: the DFS pool is 98 cohorts, of which only about 20 report a
  token literally called DFS.
* `pooling = "exact"` pools only `Exact-equivalent` rows. It is available for a
  reviewer who wants the strictest reading, and it names every cohort it leaves
  out.

Both modes record the classes they pooled, per cohort and on the pooled row.

```{r}
mk_cohort <- function(n, prefix, seed, token = "OS", role = "primary") {
  set.seed(seed)
  x <- rnorm(n, 10, 1)
  hr <- exp(0.4 * (x - 10))
  d <- data.frame(ID = sprintf("%s%03d", prefix, seq_len(n)),
                  OS_time = round(rexp(n, hr / 9) + 0.05, 3),
                  OS_status = rbinom(n, 1, 0.65),
                  GAPDH = x, TP53 = rnorm(n, 9, 1.1))
  if (token != "OS") {
    d[[paste0(token, "_time")]] <- d$OS_time
    d[[paste0(token, "_status")]] <- d$OS_status
  }
  d
}
A <- mk_cohort(200, "A", 11)                 # reports OS
B <- mk_cohort(180, "B", 12, token = "RFS")  # reports RFS (DFS family)
C <- mk_cohort(150, "C", 13, token = "DFS")  # reports DFS
merged <- list(A = A, B = B, C = C)
```

A table that marks B as related and C as exact, so the two modes differ:

```{r}
ct <- data.frame(
  Accession = c("A", "B", "C"),
  Family    = c("OS", "DFS", "DFS"),
  Token     = c("OS", "RFS", "DFS"),
  TokenRole = c("primary", "contributing", "primary"),
  PoolingClass = c("Exact-equivalent", "Clinically-related", "Exact-equivalent"),
  stringsAsFactors = FALSE)
ct
```

```{r}
strict <- cpas_meta(c("B", "C"), marker = "GAPDH", type = "DFS", merged = merged,
                    pooling = "exact", class_table = ct)
strict$per_dataset[, c("dataset", "endpoint", "pooling_class", "n", "HR")]
strict$pooled[, c("pooling", "pooling_classes", "k", "HR", "lower", "upper")]
```

## The pooling method: DL, REML and HK

`method` selects the estimator of the between-cohort variance tau^2 and the
interval of the pooled estimate:

* `"REML"` (the default since this version) - iterative restricted maximum
  likelihood;
* `"DL"` - DerSimonian-Laird, the estimator used before, kept for continuity
  (`"RE"` is accepted as a synonym);
* `"HK"` - the Hartung-Knapp-Sidik-Jonkman adjusted variance, on top of the REML
  tau^2, with its characteristically wider interval;
* `"FE"` - fixed effect.

Both prediction intervals are always returned: the primary construction selected
by `pi_method` and a clearly labelled alternative. `pi_method = "t"` (the
default) is the documented t(k-2) interval, `"normal"` is the normal
approximation and `"HK"` is the Hartung-Knapp-based interval; the rule actually
used is recorded in `$pooled$pi_method` and `$pooled$pi_rule`.

```{r}
rows <- lapply(c("DL", "REML", "HK", "FE"), function(mm) {
  p <- cpas_meta(c("A", "B", "C"), marker = "GAPDH", type = "OS", merged = merged,
                 method = mm)$pooled
  data.frame(method = mm, tau2 = p$tau2, HR = p$HR, lower = p$lower,
             upper = p$upper, p = p$p, pi_primary_lo = p$pi_lower,
             pi_alt_lo = p$pi_alt_lower)
})
do.call(rbind, rows)
```

Only the prediction interval changes with `pi_method`:

```{r}
pis <- lapply(c("t", "normal", "HK"), function(pm) {
  p <- cpas_meta(c("A", "B", "C"), marker = "GAPDH", type = "OS", merged = merged,
                 pi_method = pm)$pooled
  data.frame(pi_method = p$pi_method, HR = p$HR, pi_lower = p$pi_lower,
             pi_upper = p$pi_upper, pi_alt_lower = p$pi_alt_lower)
})
do.call(rbind, pis)
```

Leave-one-out sensitivity uses the same method as the original call:

```{r}
mt <- cpas_meta(c("A", "B", "C"), marker = "GAPDH", type = "OS", merged = merged)
loo_meta(mt)[, c("left_out", "HR", "lower", "upper", "I2")]
```

## Shared-patient overlap

Some cohort pairs share patients: one study split across two platforms, a series
deposited twice, or the same patients re-published. Pooling both members counts
those patients twice, which narrows the pooled interval without adding
information. The register of known pairs ships with the package.

```{r}
ov <- cohort_overlap()
dim(ov)
head(ov[, c("AccessionA", "AccessionB", "SharedPatients", "Basis")], 4)
```

`cpas_meta(overlap = )` acts on the pairs that are inside the pool being built:

* `"warn"` (default) proceeds, warns, and records the pairs in
  `$pooled$overlap_pairs` and in the manifest;
* `"refuse"` stops with an error naming the pair(s);
* `"dedupe"` keeps one member of every overlapping group - the larger cohort,
  ties broken by accession sort - and records what it dropped.

```{r}
ov_small <- data.frame(AccessionA = c("A", "B"), AccessionB = c("B", "C"),
                       SharedPatients = c(40L, 12L), Basis = c("title", "platform-pair"),
                       Evidence = c("identical titles", "same ids"),
                       stringsAsFactors = FALSE)
ov_small
```

```{r}
w <- cpas_meta(c("A", "B", "C"), "GAPDH", "OS", merged = merged,
               overlap_table = ov_small)
w$pooled$overlap_pairs_text
```

```{r, error = TRUE}
cpas_meta(c("A", "B", "C"), "GAPDH", "OS", merged = merged,
          overlap = "refuse", overlap_table = ov_small)
```

```{r}
dd <- cpas_meta(c("A", "B", "C"), "GAPDH", "OS", merged = merged,
                overlap = "dedupe", overlap_table = ov_small)
dd$per_dataset$dataset
dd$excluded[, c("cohort", "stage", "reason")]
```

Deduplication is deterministic: the larger cohort wins, and a tie is broken by
the accession sort, so the same pool always drops the same member.

## The analysis manifest

Everything above is recorded in one object per result, `$manifest`, also
reachable through `cpas_manifest()`. It is printable and convertible.

```{r}
cpas_manifest(mt)
```

```{r}
as.data.frame(cpas_manifest(dd))[c("field", "value")][
  as.data.frame(cpas_manifest(dd))$field %in%
    c("cohorts", "endpoint_family", "resolved_tokens_per_cohort",
      "pooling_classes_pooled", "meta_method", "tau2", "I2", "pi_primary",
      "pi_alt", "overlap_mode", "overlap_dropped"), ]
```

The structured parts stay data.frames inside the manifest, so they can be
written to CSV or compared between runs:

```{r}
str(as.data.frame(cpas_manifest(mt)), max.level = 1)
cpas_manifest(dd)$overlap_dropped
cpas_manifest(strict)$pooling_table
```

## Fail-safe: a model that would need repair is refused

The same principle applies inside a cohort. `auto_repair = FALSE` (the default)
returns an explicit `not estimable` result instead of quietly dropping a
covariate, merging a level or excluding patients; `auto_repair = TRUE` restores
the repair and records every modification.

```{r}
A$flat <- 1
bad <- COX_analysis(A, type = "OS", cont_Variates = c("GAPDH", "flat"),
                    method = "multi")
bad$status
bad$offending_terms
```

```{r}
ok <- COX_analysis(A, type = "OS", cont_Variates = c("GAPDH", "flat"),
                   method = "multi", auto_repair = TRUE)
ok$metadata$dropped_covariates[, c("variable", "reason")]
cpas_manifest(ok)$dropped_covariates
```

## Checking the annotation itself: blinded adjudication

The semantics table encodes *our* reading of each deposit, so it needs an
audit that does not simply repeat it. `endpoint_adjudication()` draws a
stratified sample (by endpoint family and by source) of cohort-endpoint records
and returns a **blinded** rating sheet: the record id, the cohort, the token and
the evidence text only - the family assignment, the pooling class and the four
semantic fields are withheld - with one empty column per judgement field and
rater.

```{r}
sheet <- endpoint_adjudication(n = 12, raters = 2, seed = 20260930)
sheet[, c("record_id", "cohort", "token")]
attr(sheet, "pooling_class_vocabulary")
```

The `record_id` is stable: it is assigned to the whole table before filtering
and sampling, so a record keeps its id between draws. Once the raters have filled
the sheet in, `endpoint_agreement()` reports the per-field raw agreement, Cohen's
kappa with its Landis-and-Koch label, the overall pooled agreement, the
adjudication rate and the disagreeing records. The example below uses a tiny
synthesised ratings frame - **not a real study result** - whose kappa is known by
hand: 20 records where both raters say `A`, 5 where only rater 1 says `A`, 10 the
other way round and 15 where both say `B`, so `po = 0.70`,
`pe = (25*30 + 25*20)/50^2 = 0.50` and `kappa = 0.40`.

```{r}
synth <- data.frame(
  record_id = sprintf("REC%04d", 1:50),
  cohort = "SYNTHETIC", token = "OS", evidence = "synthetic example",
  rater1_pooling_class = c(rep("A", 25), rep("B", 25)),
  rater2_pooling_class = c(rep("A", 20), rep("B", 5), rep("A", 10), rep("B", 15)))
ag <- endpoint_agreement(synth, fields = "pooling_class")
ag$per_field
ag$overall_raw_agreement
ag$adjudication_rate
head(ag$disagreements[, c("record_id", "fields_disagreed")], 3)
```

With more than two raters the field's kappa is the mean of the pairwise Cohen's
kappas, and the result says so.

```{r}
synth$rater3_pooling_class <- sample(c("A", "B"), 50, replace = TRUE)
ag3 <- endpoint_agreement(synth, fields = "pooling_class")
ag3$kappa_basis
ag3$pairwise[, c("pair", "n_compared", "raw_agreement", "kappa")]
```
